---
title             : "Rethinking microplastics: mapping the continuous and interrelated nature of shape, size and density of marine plastics in distributions"
shorttitle        : "Rethinking microplastics"

author: 
  - name          : "Lise Alkema"
    affiliation   : "1"
    corresponding : yes
    address       : "Postal address"
    email         : "my@email.com"
  - name          : "Caspar J. van Lissa"
    affiliation   : "2"
  - name          : "Merel Kooi"
    affiliation   : "1"
  - name          : "A. A. Koelmans"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "Wageningen University"
  - id            : "2"
    institution   : "Utrecht University, department of Methodology & Statistics"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  Methods
  
keywords          : "microplastics, plastic, maritime, mixture model"
wordcount         : "X"

bibliography      : ["r-references.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
run_everything = TRUE
library(papaja)
library(english)
source("setup.r")
if(run_everything){
  source("clean_data.r")
}
df <- read.csv("df.csv", stringsAsFactors = FALSE)

```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed,
                      echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
```

# Methods

With regard to the dimensionality of particles, we can distinguish three cases:

1. A *line* is defined as a particle that varies substantially along one dimension, and negligibly along the other two dimensions. Lines are often twisted, and therefore, it is difficult to measure their length directly. To estimate length, we divided the measured area of lines by a constant value (determined based on a sample of ...). We include a dichotomous indicator variable to distinguish lines from other particles.
2. A *film* is defined as a particle that varies substantially along two dimensions, and negligibly along the third. We have measured the Feret length and width of these particles, which describes them accurately. We include a dichotomous indicator variable to distinguish films from other particles.
3. All other particles vary substantially along three dimensions. For small particles (all dimensions < 5mm), we measured Feret length and width. For larger particles, we additionally measured height. <!--Assuming that three-dimensional particles are randomly oriented on the plane, we can assume that the correlation between the first two dimensions is similar in size to the correlation between each dimension and the third.-->

<!--Because the dimensionality of particles was assessed based on bounding boxes in two-dimensional photographs, we have to take some limitations into account. We do, however, have some information about height, -->

## Strategy of analyses

To account for the differences in particle dimensionality, we used a stepwise plan of analysis:

1. Analyze the three dimensionality categories separately, using only the dimensions relevant for that particle.
2. Determine the best model and number of classes for each dimensionality category
3. Create a joint model, with number of classes and starting values based on step 2. Use starting values for the dichotomous indicator variables to ensure lines and films are assigned to the correct classes.

### Lines

```{r, results='hide', include=FALSE}
analyzedat <- df[, c("length", "width", "Film", "Line")]

# Analyze lines ------------------------------------------------------------
analyze_line <- analyzedat[analyzedat$Line == 1, c("length")]

if(run_everything){
  res_line <- estimate_profiles(analyze_line, 1:5, variances = c("varying"), covariances = c("zero"), package = "MplusAutomation", OUTPUT = "standardized;\nsvalues;", keepfiles = TRUE)
  saveRDS(res_line, "res_line.RData")
} else {
  res_line <- readRDS("res_line.RData")
}
plot(res_line)
comp_line <- compare_solutions(res_line)

pdf("fig1.pdf")
p <- plot_density(res_line, facet_labels = c("model 2 class 1" = "1 class",
                                        "model 2 class 2" = "2 class",
                                        "model 2 class 3" = "3 class",
                                        "model 2 class 4" = "4 class",
                                        "model 2 class 5" = "5 class"))+ theme(legend.position = "none")
dev.off()
```

Because lines only vary across one dimension, we conducted univariate mixture models for 1-5 classes with varying means and variances. As indicated in Table \@ref(tab:tabclasslines1), BIC indicated that the 3-class model fitted best. However, the difference in BIC with the 2-class model was trivial, and the entropy of a 2-class solution was much higher. Most importantly, visual inspection of the solution (Figure \@ref(fig:figclasslines1)) clearly revealed a bimodal distribution. We thus chose a 2-class solution.

```{r tabclasslines1, results = "asis"}
tab_class(comp_line)
```

```{r figclasslines1, out.width="100%", fig.cap="Mixture model of lines"}
p
```

### Film

```{r, results = "hide", include = FALSE}
# Analyze film ------------------------------------------------------------
analyze_film <- analyzedat[analyzedat$Film == 1, c("length", "width")]

if(run_everything){
  res_film <- estimate_profiles(analyze_film, 1:5, variances = c("varying", "varying"), covariances = c("equal", "varying"), package = "MplusAutomation", OUTPUT = "standardized;\nsvalues;", keepfiles = TRUE)
  saveRDS(res_film, "res_film.RData")
} else {
  res_film <- readRDS("res_film.RData")
}
comp_film <- compare_solutions(res_film)
res_film

# Use plot to interpret BIC change
plot(res_film)

```
For film, we conducted bivariate mixture models for 1-5 classes. We compared models with varying means and variances, to models that also included varying covariances. As indicated in Table \@ref(tab:tabclassfilm2), BIC values hardly differed between the models with fixed, and free covariances. We thus prefer the simpler models with fixed covariances. Furthermore, BIC showed a substantial drop from the 1- to 2-class solution, followed by a smaller drop to the 3-class solution, after which the decrease stabilized. Entropy was higher for the 2-class than for the 3-class solution, indicating that the two classes were more clearly separable. Visual inspection of the solutions (Figures \@ref(fig:figclassfilm2) and \@ref(fig:figclassfilm3)) were inconclusive. Per Occam's razor, we thus retained the simpler 2-class solution.

```{r tabclassfilm2, results = "asis"}
tab_class(comp_film)
```

```{r figclassfilm2, out.width="100%", fig.cap="Mixture model of film with free means and variances, and fixed covariances"}
plot_bivariate(res_film$model_4_class_2)
```

```{r figclassfilm3, out.width="100%", fig.cap="Mixture model of film with free means and variances, and fixed covariances"}
plot_bivariate(res_film$model_4_class_3)
```

### Other

```{r, results = "hide", include=FALSE}
analyze_other <- df[!(analyzedat$Film == 1|analyzedat$Line == 1), c("length", "width")]

if(run_everything){
  res_other <- estimate_profiles(analyze_other, 1:5, variances = c("varying", "varying"), covariances = c("equal", "varying"), package = "MplusAutomation", OUTPUT = "standardized;\nsvalues;", keepfiles = TRUE)
  saveRDS(res_other, "res_other.RData")
} else {
  res_other <- readRDS("res_other.RData")
}
res_other
plot(res_other)
comp_other <- compare_solutions(res_other)
```

For the remaining particles, we conducted the same bivariate mixture models as for film. Similar to the models for film, BIC values hardly differed between the models with fixed, and free covariances, Table \@ref(tab:tabclassother3). We thus prefer the simpler models with fixed covariances. Furthermore, BIC showed a substantial drop from the 1- to 2-class solution, after which the decrease stabilized. Entropy was higher for the 2-class than for the 3-class solution, indicating that the two classes were more clearly separable. Visual inspection of the solutions (Figures \@ref(fig:figclassother4) and \@ref(fig:figclassother5)) were inconclusive. Per Occam's razor, we thus retained the simpler 2-class solution. 


```{r tabclassother3, results = "asis"}
tab_class(comp_other)
```

```{r figclassother4, out.width="100%", fig.cap="Mixture model of other particles with free means and variances, and fixed covariances"}
plot_bivariate(res_other$model_4_class_2)
```

```{r figclassother5, out.width="100%", fig.cap="Mixture model of other particles with free means and variances, and fixed covariances"}
plot_bivariate(res_other$model_4_class_3)
```

## Combining the classes

```{r, results = "hide", include = FALSE}
df_combined <- df
df_combined[paste0("c", 1:6)] <- 0
df_combined[df_combined$Line == 1, paste0("c", 1:2)] <- 1
df_combined$width[df_combined$category == "Line"] <- NA
df_combined[df_combined$Film == 1, paste0("c", 3:4)] <- 1
df_combined[!(df_combined$Line == 1 | df_combined$Film == 1), paste0("c", 5:6)] <- 1
df_combined <- df_combined[, c("sample", "length", "width", "height_obs", paste0("c", 1:6))]
if(run_everything){
  model <- mplusObject(TITLE = "Combined model",
                       VARIABLE = "USEVARIABLES = length width c1 c2 c3 c4 c5 c6;\nCLASSES = c(6);\nTRAINING = c1 c2 c3 c4 c5 c6;\nCLUSTER = sample;",
                       ANALYSIS = c("
                                    TYPE = complex mixture;",
                                    "STARTS = 100 10;",
                                    "STITERATIONS = 20;"),
                       MODEL = c("%c#1%",
                                 "length (vlength1);",
                                 "width (vwidth3);",
                                 "[ length*14.07023 ];",
                                 "[width@0];",
                                 "length WITH width (lengthWwidth3);",
                                 "%c#2%",
                                 "length (vlength2);",
                                 "width (vwidth3);",
                                 "[ length*48.96305 ];",
                                 "[width@0];",
                                 "length WITH width (lengthWwidth3);",
                                 "%c#3%",
                                 "length (vlength3);",
                                 "width (vwidth3);",
                                 "[ length*3.06779 ];",
                                 "[ width*1.89682 ];",
                                 "length WITH width (lengthWwidth3);",
                                 "%c#4%",
                                 "length (vlength4);",
                                 "width (vwidth4);",
                                 "[ length*9.29501 ];",
                                 "[ width*5.63437 ];",
                                 "length WITH width (lengthWwidth3);",
                                 "%c#5%",
                                 "length (vlength5);",
                                 "width (vwidth5);",
                                 "[ length*2.20399 ];",
                                 "[ width*1.56727 ];",
                                 "length WITH width (lengthWwidth5);",
                                 "%c#6%",
                                 "length (vlength6);",
                                 "width (vwidth6);",
                                 "[ length*5.10412 ];",
                                 "[ width*3.40502 ];",
                                 "length WITH width (lengthWwidth5);"), 
                       OUTPUT = c("TECH14;",
                                  "standardized;",
                                  "svalues;"),
                       SAVEDATA = c("FILE IS combined_save.dat;",
                                    "SAVE = cprobabilities;"),
                       rdata = df_combined)
  manual_combined <- mplusModeler(model, modelout = "combined.inp", run = 1L)$results
} else {
  manual_combined <- readModels("combined.out")
}
res <- tidyLPA:::as.tidyLPA(manual_combined)
```

To combine the models for lines, films, and other particles, we estimated a six-class mixture model. We used dummy variables to restrict potential class membership for the three categories to two classes each. Parameters involving width were not estimated for line classes. To account for non-independence of observations due to the clustered sampling, we used a sandwich estimator for the standard errors.

The resulting model discriminated well between classes, Entropy $`r report(res[["model_unknown_model_1_class_6"]][["fit"]]["Entropy"])`$, posterior classification probabilities $[`r report(res[["model_unknown_model_1_class_6"]][["fit"]]["prob_min"], equals = FALSE)`, `r report(res[["model_unknown_model_1_class_6"]][["fit"]]["prob_max"], equals = FALSE)`]$, AIC `r report(res[["model_unknown_model_1_class_6"]][["fit"]]["AIC"])`, BIC `r report(res[["model_unknown_model_1_class_6"]][["fit"]]["BIC"])`. The parameters of this classification are displayed in Tables \@ref(tab:tabfinal1)-\@ref(tab:tabfinal3).

```{r tabfinal1, results = "asis"}
tab <- manual_combined$parameters$unstandardized
tabstd <- manual_combined$parameters$stdyx.standardized
#tab <- readRDS("c:/git_repositories/lise_paper/tab.RData")
tab <- tab[tab$param %in% c("LENGTH", "WIDTH"), ]
tabstd <- tabstd[tabstd$param %in% c("LENGTH", "WIDTH"), ]
tab[grepl("WITH$", tab$paramHeader), c("est", "se", "pval")] <- 
  tabstd[grepl("WITH$", tabstd$paramHeader), c("est", "se", "pval")]
tab$CI <- confint(tab$est, se = tab$se)
tab$label <- paste0(tab$paramHeader, ".", tab$param)
tab <- tab[, c(9, 3, 8,7)]
tmp <- reshape(tab, v.names = c("est", "CI"), timevar = "LatentClass", idvar = "label", direction = "wide")
names(tmp)<- c("Parameter", rep(c("Est", "CI"), 6))
tmp$Parameter <- c("$r_{\\text{length}, \\text{width}}$", "$M_{\\text{length}}$", "$M_{\\text{width}}$", "$s^2_{\\text{length}}$", "$s^2_{\\text{width}}$")

colspans <- mapply(function(x, y){c(x, y)}, x = seq(from = 2, to = 13, by = 2), y = seq(from = 3, to = 13, by = 2), SIMPLIFY = FALSE)
names(colspans) <- paste0("Class ", 1:6, " (N = ", table(manual_combined$savedata$C), ")")
tb <- tmp[c(2,4), 1:5]
names(tb)[4:5] <- names(tb)[2:3]
rownames(tb) <- NULL
apa_table(tb, col_spanners = colspans[1:2], escape = FALSE, format.args = list(na_string = ""))
```

```{r tabfinal2, results = "asis"}
colspans2 <- list(c(2,3), c(4,5))
names(colspans2) <- names(colspans)[3:4]
tb <- tmp[, c(1, 6:9)]
names(tb)[4:5] <- names(tb)[2:3]
rownames(tb) <- NULL
apa_table(tb, col_spanners = colspans2, escape = FALSE,format.args = list(na_string = ""))
```
```{r tabfinal3, results = "asis"}
colspans3 <- list(c(2,3), c(4,5))
names(colspans3) <- names(colspans)[5:6]
tb <- tmp[, c(1, 10:13)]
names(tb)[4:5] <- names(tb)[2:3]
rownames(tb) <- NULL
apa_table(tb, col_spanners = colspans3, escape = FALSE,  format.args = list(na_string = ""))
```


## Observed height

```{r, echo = FALSE}
df_height <- manual_combined$savedata
df_height <- cbind(Height = df$height_obs, df_height)
tab_height <- table(df_height$C, is.na(df_height$Height))
height_cors <- cor(df_height[df_height$C == 6, c("LENGTH", "WIDTH", "Height")], use = "pairwise.complete.obs")
```
Observed height was only assessed for particles with one dimension exceeding 5mm. Consequently, we did not involve it in the classification procedure, but we do explore relationships with observed height as a distal outcome. First, we examined the prevalence of observed height measurements across the six particle classes, based on most likely class membership. We found that only `r as.english(sum(tab_height[1:5, 1]))` particles with observed height were not classified as *large fragments*. Consequently, we procede to analyze observed height only for particles classified as *large fragments*. For these particles, we found that length was uncorrelated with height, $r `r report(height_cors[1,3])`$, and width had a small correlation with height, $r `r report(height_cors[2,3])`$. This again reinforces the notion that larger particles display greater heterogeneity in shapes.

\newpage

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id = "refs"></div>
\endgroup
